---
title: "FIA-Post-Hurricane-Assessment"
author: "Ben Branoff & Andrés Baeza-Castro"
date: "2026-08-10"
output: github_document
always_allow_html: true
---
```{r load-packages,warning=FALSE,message=FALSE,echo=FALSE}
library(terra)
library(Cyclones)
library(sf)
library(dplyr)
library(knitr)
library(kableExtra)
```
# FIA-Post-Hurricane-Assessment
Hurricane damage/loss estimates from collected plots after storms


Start by loading the plots from nims_srs_db.   Here, using the DBI library, which requires some initial setup for successful credentialed connections to the db.
Once setup, can connect to the "fiadb01p" data base and access any tables you are authorized to access. Then, we can use SQL queries passed as string 
arguments to retrieve the desired data. Here we are asking for the Florida (132024, 132025) and Georgia (122024, 122025) evaluations since hurricane Helene.
For now, we are only asking for the plots, their conditions, and their associated strata variables. Their trees and reference tables will be joined later.  
```{r load-plots,cache=TRUE,warning=FALSE,message=FALSE}
library(DBI)
library(sf)
library(knitr)
library(kableExtra)
myconn <- dbConnect(odbc::odbc(), "fiadb01p", timeout = 10) # function change w/DBI
vars = paste0("select p.cn,p.prev_plt_cn,p.plot_status_cd,p.invyr,p.eval_grp,p.statecd,p.countycd,p.measyear,p.measmon,p.measday,pa.actual_lat,pa.actual_lon,",
                    "c.condid,c.cond_status_cd,c.owncd,c.fortypcd,c.condprop_unadj,c.prop_basis,",
              "ps.evalid,ps.estn_unit_cn,ps.adj_factor_macr,ps.adj_factor_subp,ps.expns,ps.P1POINTCNT,ps.P2POINTCNT,",
              "pps.stratum_cn,pc.totplots,pc2.nplots,",
              "PEU.p1pntcnt_eu,PEU.area_used ")
tabs = "from fs_nims_fiadb_srs.plotsnap p,fs_nims_fiadb_srs.sds_plot pa,fs_nims_fiadb_srs.cond c,fs_nims_fiadb_srs.pop_stratum ps,fs_nims_fiadb_srs.pop_plot_stratum_assgn pps,(select evalid,count(cn) as totplots from fs_nims_fiadb_srs.pop_plot_stratum_assgn where evalid in (122401,132401,122400,132400) group by evalid) pc,(select stratum_cn,count(cn) as nplots from fs_nims_fiadb_srs.pop_plot_stratum_assgn where evalid in (122401,132401,122400,132400) group by stratum_cn) pc2,fs_nims_fiadb_srs.POP_ESTN_UNIT PEU "
filts = paste0("where p.eval_grp in (122024,132024) 
and ps.evalid in (122401,132401,122400,132400)
and c.plt_cn = p.cn
and pa.plt_cn = p.cn
and pps.stratum_cn = ps.cn
and pps.plt_cn = p.cn
and pc.evalid=pps.evalid
and pc2.stratum_cn=ps.cn
and PEU.CN = PS.ESTN_UNIT_CN")
#and to_date(p.measyear || '-' || LPAD(p.measmon, 2, '0') || '-' || LPAD(p.measday, 2, '0')
#               default null on conversion error,
#               'YYYY-MM-DD') >= date '2024-09-26'") 
plots = dbGetQuery(myconn,paste0(vars,tabs,filts))
##  also, convert the plots to a spatial object
plots_geo <- st_as_sf(plots,coords=c("ACTUAL_LON","ACTUAL_LAT"),crs=4326)
plots |> head() |>
  kable(format='html')|>
  kable_styling() |>
  scroll_box(width='100%')
```

The above is a small subset, but there are around 8000 plots that have been collected since the storm in Georgia and Florida.

Now, use the Cyclones package to recreate the winds for Helene and create wind intensity zones. For now, Cyclones is only available on
Github, so an extra package ("remotes") is required to install it. For this workflow, the hurricane wind information is provided as part of the download and there is no need to recalculate, but the process is demonstrated anyways.

Below demonstrates how the above hurricane data was obtained. First the Cyclones package is downloaded and installed and it is used to ingest the necessary data and process into the wind fields we need.  
First, download the package and get the storm track information. This is the foundation for being able to retrieve the winds (and rain and storm surge).
```{r load-winds,eval=TRUE,cache=TRUE,message=FALSE,warning=FALSE}
##  install, only needed once
# install.packages('remotes')
# remotes::install_github("BBranoff/Cyclones@development")

library(Cyclones)
##  get the storm information
Helene <- get_storms(name="HELENE",season=2024,erddap=FALSE)
Helene$HELENE_2024_NA_2024268N17278
```

Now, calculate the wind field. Here we just use a theoretical (Boose) equation based on the maximum wind speed.
We calculate the wind onto a grid with a spatial resolution of 5km.

```{r winds,eval=TRUE,cache=TRUE,message=FALSE,warning=FALSE,fig.width=9,fig.height=5,fig.cap = "Fig 1. Hurricane Helene's maximum sustained winds in meters per second."}
###  first, interpolate the spatial wind information to every 30 mins
Helene_extents <- make_extents(Helene,t_res=30)
###  then interpolate spatially as a raster
Helene_winds <- get_wind(Helene_extents,method="Boose",agg=TRUE,s_res=5000)

##  get some boundaries, mostly for display
##  county boundaries
counties <- read_sf("/vsizip//vsicurl/https://naciscdn.org/naturalearth/10m/cultural/ne_10m_admin_2_counties.zip",
                        layer = "ne_10m_admin_2_counties")
usa <- counties|>filter(REGION_COD %in% c(12,13,47,45,37))|>group_by(REGION_COD)|>summarise()
###  The FIA state boundary is important because it determines the total size of the estimation unit
###  we use this to calculate the expansion factor
##   normally this is on the T drive but its copied to Github for easier access
FIAstates <- read_sf("https://github.com/BBranoff/FIA-Post-Hurricane-Assessment/raw/refs/heads/main/data/rs_states.gpkg")

library(terra)
plot(Helene_winds$boose$max_msw,main="Maximum Sustained Winds (m/s)")
plot(st_geometry(st_transform(usa,crs(Helene_winds$boose$max_msw))),add=TRUE)
```

We can then categorize the wind into the familiar Saffir-Simpson scale. These categories will be our intensity zones and they will be the on-the-fly strata that we use to calculate sub-complete evaluation estimates. 

```{r load-raster,warning=FALSE,message=FALSE,echo=FALSE}
##  caching doesnt work on raster objects, because they are held in memory
Helene_winds <- terra::unwrap(readRDS(gzcon(url("https://github.com/BBranoff/FIA-Post-Hurricane-Assessment/raw/refs/heads/main/data/Helene.RDS")))$winds$boose)
###  the EVALIDATOR API functions, also stored on Github for easy access
source("https://github.com/BBranoff/FIA-Post-Hurricane-Assessment/raw/refs/heads/main/data/EVALIDATOR%20API.R")
FIAstates <- read_sf("https://github.com/BBranoff/FIA-Post-Hurricane-Assessment/raw/refs/heads/main/data/rs_states.gpkg")
counties <- read_sf("/vsizip//vsicurl/https://naciscdn.org/naturalearth/10m/cultural/ne_10m_admin_2_counties.zip",
                        layer = "ne_10m_admin_2_counties")
usa <- counties|>filter(REGION_COD %in% c(12,13,47,45,37))|>group_by(REGION_COD)|>summarise()
```
```{r winds-hidden,eval=TRUE,fig.width=9,fig.height=5,fig.cap = "Fig 2. Hurricane Helene's wind field, categorized into intensity zones." }
###  classify the maximum wind speed into the Saffir-Simpson scale
Helene_Cat <- as.factor(classify(Helene_winds$max_msw,
                       matrix(c(0,33,0,
                                33,43,1,
                                43,50,2,
                                50,60,3,
                                60,70,4,
                                70,Inf,5),ncol=3,byrow=TRUE),include.lowest=TRUE))
plot(Helene_Cat,main="Wind Intensity Categories")
plot(st_geometry(st_transform(usa,crs(Helene_Cat))),add=TRUE)
```

The Helene hurricane data is now processed and ready for overlays with the plots. 

With the hurricane intensity areas defined, we can assign them to each plot. We also need to define the area of each wind zone. 
These will be our artificial 'strata', and we will need to know the total area in order to calculate the expansion factors. This is done two ways. The first is using vectors (shapefiles) of the strata and the 'st_area()' function to calculate the areas. The second is using the raster image and counting the number of pixels in each strata, then multiplying by the resolution to get total area. 

```{r extract,eval=TRUE,fig.width=9,fig.height=9,fig.cap = "Fig 3. Hurricane Helene's wind zones with the plots overlayed. Plots are assigned their respective wind zones and their expansion factors come from the total area of each wind zone divided by the number of plots in each zone." }
###  first, mask out the areas outside of our interest
Helene_Cat <- trim(mask(Helene_Cat,st_transform(FIAstates|>filter(STATEFP %in% c(12,13)),crs(Helene_Cat))))
#Helene_Cat100 <- project(Helene_Cat,rast(ext=ext(Helene_Cat),crs=crs(Helene_Cat),res=c(100,100)))
# Helene_Cat_vect <- st_as_sf(as.polygons(Helene_Cat))|>st_intersection(FIAstates|>filter(STATEFP %in% c(12,13))|>st_transform(crs(Helene_Cat)))%>%
#   mutate(VectArea.m = st_area(.),
#          VectArea.ac=VectArea.m/4046.86)|>
#   group_by(STATEFP)|>
#   mutate(statearea.ac=sum(VectArea.ac))

##  the raster area approach to pixel counting
##  important to do this while the raster is in projected, not geographic, coordinates
##  also important to do by state, as this is how the evaluation is done
# Areas12 <- table(values(mask(Helene_Cat,st_transform(FIAstates|>filter(STATEFP==12),crs(Helene_Cat)))))
# Areas13 <- table(values(mask(Helene_Cat,st_transform(FIAstates|>filter(STATEFP==13),crs(Helene_Cat)))))
# Areas <- rbind(data.frame(Areas12)|>mutate(STATECD=12),
#                data.frame(Areas13)|>mutate(STATECD=13)) |>
#   rename(npixels=Freq)|>
#   mutate(area.raster.m = npixels*res(Helene_Cat)[1]*res(Helene_Cat)[2],
#          area.raster.ac = area.raster.m/4046.86)|>
#   group_by(STATECD)|>
#   mutate(state_tot_pixels=sum(npixels),
#          state_tot_pixels_area = sum(area.raster.ac))|>
#   left_join(Helene_Cat_vect|>
#               st_drop_geometry()|>
#               mutate(STATECD=as.numeric(STATEFP)),by=join_by(STATECD,Var1==max_msw))|>
#   st_drop_geometry()|>ungroup()|>
#   ##  compare to FIA state total areas from the ESTN_UNITs
#   left_join(plots |> 
#               group_by(EVALID,EVAL_GRP)|>
#               mutate(totFIAplots=length(unique(CN)))|>
#               distinct(ESTN_UNIT_CN,.keep_all=TRUE)|>
#               group_by(EVAL_GRP,EVALID,totFIAplots)|>summarise(totFIAarea=sum(AREA_USED))|>
#               filter((EVAL_GRP ==122024&EVALID==122401)|(EVAL_GRP==122025&EVALID==122501)|(EVAL_GRP==132025&EVALID==132501)|(EVAL_GRP==132024&EVALID==132401))|>
#               mutate(STATECD=as.numeric(substr(EVALID,1,2)))|>ungroup()|>distinct(STATECD,totFIAarea,totFIAplots),
#             by=join_by(STATECD))
# 
# Areas

# ###  check with the units on file
# FIAunits<- read_sf("//usda.net/fs/FS/RD/SRS/Science/FIA/Collaboration/SRS_GIS_Layers/SRS/srs_unit.shp") |>
#   filter(WU >100 & WU<140)|>
#   mutate(STATECD = substr(WU,1,2))
# ###  need the national forests as well
# forests <- read_sf("C:/Users/BenjaminBranoff/Downloads/BdyAdm_LSRS_AdministrativeForest/S_USA.BdyAdm_LSRS_AdministrativeForest.shp")|> 
#   filter(FORESTNAME %in% c("Chattahoochee-Oconee National Forests","National Forests in Florida"))
# ###  combine for all ESTIMATION UNITS in analysis
# FLGAunits <- rbind(st_difference(FIAunits |> rename(Unit=WU)|>select(Unit,STATECD),
#                    st_union(forests|> rename(Unit=FORESTNAME)|>select(Unit,STATECD))),
#                    forests|> rename(Unit=FORESTNAME)|>select(Unit,STATECD))|>
#   ## calculate area
#   st_transform(crs(Helene_Cat)) |>
#   st_as_sf() %>%
#   mutate(EU_Area.ac = as.numeric(st_area(.))/4046.86,
#          EU_nPixels=zonal(Helene_Cat,vect(.),fun="notNA")$max_msw)
# 
# dist_threshold <- 10000 # e.g., 1000 meters
# strataPixels <- lapply(unique(plots_geo$STRATUM_CN),function(x){
#  r= plots_geo |> filter(STRATUM_CN==x) |>
#     st_transform(crs(Helene_Cat))%>%
#     rasterize(field='STRATUM_CN',Helene_Cat500)
#   sum(!is.na(values(r)))
# })
# EU_pixels <- sum(!is.na(values(plots_geo |>
#     st_transform(crs(Helene_Cat))%>%
#     rasterize(field='STRATUM_CN',Helene_Cat500))))
# strata <-plots_geo |> 
#  distinct(STRATUM_CN)|>
#   mutate(nPixels=unlist(strataPixels))
# 
# strata <- plots_geo |> group_by(STRATUM_CN)|>
#   summarise(do_union=TRUE) %>%
#   st_concave_hull(.,ratio=0.01)|>
#   st_as_sf() %>%
#   st_transform(crs(Helene_Cat))%>%
#   mutate(area.ac=as.numeric(st_area(.))/4046.86)%>%
#   mutate(nPixels=zonal(Helene_Cat,vect(.),fun="notNA")$max_msw)
# unitPixels <- lapply(unique(plots_geo$ESTN_UNIT_CN),function(x){
#  r= plots_geo |> filter(ESTN_UNIT_CN==x) |>
#     st_transform(crs(Helene_Cat))%>%
#     rasterize(field='ESTN_UNIT_CN',Helene_Cat500)
#   sum(!is.na(values(r)))
# })
# units<-plots_geo |> 
#  distinct(ESTN_UNIT_CN)|>
#   mutate(nPixels_EU=unlist(unitPixels))
# 
# units <- plots_geo |> group_by(ESTN_UNIT_CN)|>
#   summarise(do_union=TRUE) %>%
#   st_concave_hull(.,ratio=0.5)|>
#   st_as_sf() %>%
#   st_transform(crs(Helene_Cat))%>%
#   mutate(area_EU.ac=as.numeric(st_area(.))/4046.86)%>%
#   mutate(nPixels_EU=zonal(Helene_Cat,vect(.),fun="notNA")$max_msw)


####  the weighting for the EXPNS calculation can be approximated closesly by the 
####  plot ratio, instead of the pixel ratio that the offical calculation uses
plots_geo <- plots_geo |>
  mutate(PixelRatio=P1POINTCNT/P1PNTCNT_EU)|>
  group_by(ESTN_UNIT_CN)|>
  mutate(ESTN_UNIT_nPlots = sum(P2POINTCNT[!duplicated(STRATUM_CN)]))|>
  ungroup()|>
  #left_join(strata|>st_drop_geometry(),by=join_by(STRATUM_CN))|>
  #left_join(units|>st_drop_geometry(),by=join_by(ESTN_UNIT_CN))|>
  group_by(STRATUM_CN)|>
  mutate(nplots=length(unique(CN)),
         plotratio = nplots/ESTN_UNIT_nPlots, 
         EXPNS2 = plotratio*(AREA_USED/nplots))

plot(plots_geo$PixelRatio,plots_geo$plotratio)
abline(a=0,b=1,lty=2)
###  this is the key to approximating the EXPNS
plot(plots_geo$EXPNS,plots_geo$EXPNS2)
abline(a=0,b=1,lty=2)

###  project onto geographic coordinates
Helene_Cat <- project(Helene_Cat,"epsg:4326")
plot(Helene_Cat)
###  overlay the plots
plot(st_geometry(st_as_sf(plots_geo,coords=c("LON","LAT"),crs=4326)),add=TRUE,pch=19,cex=0.01,col="darkgrey")
plot(st_geometry(usa),add=TRUE,border="darkgrey")
plot(st_geometry(counties),add=TRUE,border="darkgrey")
```

We now have the plots and the areas associated with each wind zone. From here, we can begin to assemble the
necessary tables and calculate the desired estimates. 

Beginning with assembling the necessary tables. Forested area is likely the most straight forward. We will get the 
forested acreage of each plot and, using a custom expansion factor, estimate the total forested acreage in each wind zone.

First, we need some helper functions to compare our estimates with 'official' numbers from FIA. We use the FIA EVALIDATOR API to do this. These functions are provided here by following the below steps, but they are taken directly from the [FIA EVALIDATOR API instructions](https://apps.fs.usda.gov/fiadb-api/). 
```{r evalidator_f,eval=FALSE,echo=FALSE}
##  to load the functions necessary for the EVALIDATOR API
##  source them directly from the Github repository for this package
##  They can also be copied from: https://apps.fs.usda.gov/fiadb-api/
source("https://github.com/BBranoff/FIA-Post-Hurricane-Assessment/raw/refs/heads/main/data/EVALIDATOR%20API.R")
```

Now, extract the wind zones at each plot, calculate the new expansion factor based on the area of each strata (wind zone) and then compute the forested area for each zone.  

As a proof of concept, we use the original expansion factor (EXPNS) and sum everything within an EVAL_GRP. This should be equivalent to what is computed by EVALIDATOR. To check, we use the API to retrieve the EVALIDATOR estimate and compare it to ours. They are identical, meaning we are calculating correctly for a full evaluation. We then try the same calculation using EXPNS_new to estimate the areas for the custom wind zones. 
```{r evalidate,eval=TRUE}
###  first, join the total 'strata' (wind zone) areas to the plot table
plots_geo$WindZone <- extract(Helene_Cat,plots_geo)$max_msw

###  try first regardless of windzone to test concept
###  simulating the SQL query from ref_pop_attribute
plot_summary_cust <- plots_geo |>
  st_drop_geometry()|>
  filter(EVAL_GRP %in% c(122024,132024),EVALID %in% c(122401,132401),
         COND_STATUS_CD == 1,!is.null(CONDPROP_UNADJ))|>
  mutate(Y_HID_ADJUSTED = if_else(PROP_BASIS=="MACR",CONDPROP_UNADJ*ADJ_FACTOR_MACR,CONDPROP_UNADJ*ADJ_FACTOR_SUBP))|>
  group_by(EVAL_GRP,EVALID,ESTN_UNIT_CN, STRATUM_CN, CN)|>
  summarise(Y_HID_ADJUSTED=sum(Y_HID_ADJUSTED))|>
  ungroup()
phase_summary_cust <- plot_summary_cust |>
  group_by(EVAL_GRP,STRATUM_CN, ESTN_UNIT_CN, EVALID)|>
  summarise(YSUM_HD=sum(Y_HID_ADJUSTED),
          YSUM_HD_SQR=sum(Y_HID_ADJUSTED*Y_HID_ADJUSTED),
          plots_in_domain=n(),
          non_zero_plots=sum(!is.null(Y_HID_ADJUSTED)|Y_HID_ADJUSTED>0))
phase1_summary_cust <- plots_geo |>
  st_drop_geometry()|>
  distinct(STRATUM_CN,.keep_all=TRUE)|>
  group_by(EVAL_GRP,EVALID,ESTN_UNIT_CN)|>
  mutate( W_H=P1POINTCNT/sum(P1POINTCNT),
          total_area=unique(AREA_USED),
          N=sum(P2POINTCNT),
          N_H=P2POINTCNT)|>
  ungroup()
phase_join_cust <- phase1_summary_cust |>
  left_join(phase_summary_cust,by=join_by(STRATUM_CN,EVAL_GRP,EVALID,ESTN_UNIT_CN))
estimate <-phase_join_cust |>
  mutate(YSUM_HD=if_else(is.na(YSUM_HD),0,YSUM_HD),
         YSUM_HD_SQR=if_else(is.na(YSUM_HD_SQR),0,YSUM_HD_SQR)) |>
  group_by(EVAL_GRP,EVALID,ESTN_UNIT_CN,total_area,N) |>
  summarise(estimate=sum(YSUM_HD*EXPNS),
            estimate_cust=sum(YSUM_HD*EXPNS2),
            total_plots=unique(N),
            domain_plots=sum(plots_in_domain,na.rm=TRUE),
            non_zero_plots=sum(non_zero_plots,na.rm=TRUE),
            var_of_estimate = unique(total_area) * unique(total_area) / total_plots*
               ((sum(W_H * N_H *
                     (((YSUM_HD_SQR / N_H) -
                     ((YSUM_HD / N_H) *
                     ((YSUM_HD / N_H)))) /
                     (N_H - 1)))) +
               1 / total_plots *
               (sum((1 - W_H) * N_H *
                     (((YSUM_HD_SQR / N_H) -
                     ((YSUM_HD / N_H) *
                     (YSUM_HD / N_H))) /
                     (N_H - 1))))))|>
  ungroup()

###  our custom estimate versus EVALIDATOR
EVALID_122024 <- fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=122024&outputFormat=NJSON")
EVALID_132024 <- fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=132024&outputFormat=NJSON")

estimate |> filter(EVAL_GRP %in% c(122024,132024),EVALID %in% c(122401,132401))|>
  group_by(EVAL_GRP)|>
  summarise(estimate_orig=sum(estimate),
          estimate_cust=sum(estimate_cust),
          variance_orig=sum(var_of_estimate))|>
  mutate(EVALIDATORacres = as.numeric(c(EVALID_122024$totals$ESTIMATE,EVALID_132024$totals$ESTIMATE)),
         EVALIDATORvariance= as.numeric(c(EVALID_122024$totals$VARIANCE,EVALID_132024$totals$VARIANCE)))|>
  kable(format='html')|>
  kable_styling() |>
  scroll_box(width='100%')

  

```

This is equivalent to the following calls directly to the DB>

```{r db_est,cache=TRUE,warning=FALSE,message=FALSE,eval=FALSE}
library(DBI)
library(sf)
library(knitr)
library(kableExtra)
phase1_summ <- "SELECT PEV.CN EVAL_CN,
                       PEG.EVAL_GRP,
                       PEG.EVAL_GRP_DESCR,
                       PEG.CN POP_EVAL_GRP_CN,
                       POP_STRATUM.ESTN_UNIT_CN,
                       POP_STRATUM.EXPNS,
                       POP_STRATUM.CN POP_STRATUM_CN,
                       P1POINTCNT /
                        (SELECT SUM(STR.P1POINTCNT)
                          FROM fs_nims_fiadb_srs.POP_STRATUM STR
                         WHERE STR.ESTN_UNIT_CN = POP_STRATUM.ESTN_UNIT_CN) W_H,
                       (SELECT SUM(STR.P1POINTCNT)
                          FROM fs_nims_fiadb_srs.POP_STRATUM STR
                         WHERE STR.ESTN_UNIT_CN = POP_STRATUM.ESTN_UNIT_CN) N_PRIME,
                       P1POINTCNT N_PRIME_H,
                       (SELECT SUM(EU_S.AREA_USED)
                          FROM fs_nims_fiadb_srs.POP_ESTN_UNIT EU_S
                         WHERE EU_S.CN = POP_STRATUM.ESTN_UNIT_CN) TOTAL_AREA,
                       (SELECT SUM(STR.P2POINTCNT)
                          FROM fs_nims_fiadb_srs.POP_STRATUM STR
                         WHERE STR.ESTN_UNIT_CN = POP_STRATUM.ESTN_UNIT_CN) N,
                       POP_STRATUM.P2POINTCNT N_H
                  FROM fs_nims_fiadb_srs.POP_EVAL_GRP PEG
                  JOIN fs_nims_fiadb_srs.POP_EVAL_TYP PET
                    ON (PET.EVAL_GRP_CN = PEG.CN)
                  JOIN fs_nims_fiadb_srs.POP_EVAL PEV
                    ON (PEV.CN = PET.EVAL_CN)
                  JOIN fs_nims_fiadb_srs.POP_ESTN_UNIT PEU
                    ON (PEV.CN = PEU.EVAL_CN)
                  JOIN fs_nims_fiadb_srs.POP_STRATUM POP_STRATUM
                    ON (PEU.CN = POP_STRATUM.ESTN_UNIT_CN)
                 WHERE PEG.EVAL_GRP IN (122024)
                   AND PET.EVAL_TYP = 'EXPCURR'"
phase1_summ <- dbGetQuery(myconn,gsub("\\s+", " ", phase1_summ)) 
phase_summary <- "SELECT POP_STRATUM_CN,
       ESTN_UNIT_CN,
       EVAL_CN,
       SUM(Y_HID_ADJUSTED) YSUM_HD,
       SUM(Y_HID_ADJUSTED * Y_HID_ADJUSTED) YSUM_HD_SQR,
       COUNT(*) NUMBER_PLOTS_IN_DOMAIN,
       SUM(CASE
             WHEN Y_HID_ADJUSTED IS NULL THEN
              0
             WHEN Y_HID_ADJUSTED = 0 THEN
              0
             ELSE
              1
           END) NON_ZERO_PLOTS
  FROM (SELECT 2 ATTRIBUTE_NBR,
               SUM(COND.CONDPROP_UNADJ * CASE COND.PROP_BASIS
                     WHEN 'MACR' THEN
                      POP_STRATUM.ADJ_FACTOR_MACR
                     ELSE
                      POP_STRATUM.ADJ_FACTOR_SUBP
                   END) AS Y_HID_ADJUSTED,
               PEU.CN ESTN_UNIT_CN,
               PEV.CN EVAL_CN,
               POP_STRATUM.CN POP_STRATUM_CN,
               PLOT.CN PLT_CN
          FROM fs_nims_fiadb_srs.POP_EVAL_GRP PEG
          JOIN fs_nims_fiadb_srs.POP_EVAL_TYP PET
            ON (PET.EVAL_GRP_CN = PEG.CN)
          JOIN fs_nims_fiadb_srs.POP_EVAL PEV
            ON (PEV.CN = PET.EVAL_CN)
          JOIN fs_nims_fiadb_srs.POP_ESTN_UNIT PEU
            ON (PEV.CN = PEU.EVAL_CN)
          JOIN fs_nims_fiadb_srs.POP_STRATUM POP_STRATUM
            ON (PEU.CN = POP_STRATUM.ESTN_UNIT_CN)
          JOIN fs_nims_fiadb_srs.POP_PLOT_STRATUM_ASSGN POP_PLOT_STRATUM_ASSGN
            ON (POP_PLOT_STRATUM_ASSGN.STRATUM_CN = POP_STRATUM.CN)
          JOIN fs_nims_fiadb_srs.PLOT
            ON (POP_PLOT_STRATUM_ASSGN.PLT_CN = PLOT.CN)
          JOIN fs_nims_fiadb_srs.PLOTGEOM
            ON (PLOT.CN = PLOTGEOM.CN)
          JOIN fs_nims_fiadb_srs.COND
            ON (COND.PLT_CN = PLOT.CN)
         WHERE COND.COND_STATUS_CD = 1
           AND COND.CONDPROP_UNADJ IS NOT NULL
           AND PET.EVAL_TYP = 'EXPCURR'
           AND PEG.EVAL_GRP IN (122024)
           AND 1 = 1
         GROUP BY PEU.CN, PEV.CN, POP_STRATUM.CN, PLOT.CN) PLOT_SUMMARY
 GROUP BY POP_STRATUM_CN, ESTN_UNIT_CN, EVAL_CN"
phase_summary <- dbGetQuery(myconn,gsub("\\s+", " ",phase_summary)) 
phase_join <- phase1_summ |>
  left_join(phase_summary,by=join_by(POP_STRATUM_CN,EVAL_CN,ESTN_UNIT_CN))
phase_join <-phase_join |>
  mutate(YSUM_HD=if_else(is.na(YSUM_HD),0,YSUM_HD),
         YSUM_HD_SQR=if_else(is.na(YSUM_HD_SQR),0,YSUM_HD_SQR)) |>
  group_by(POP_EVAL_GRP_CN,EVAL_GRP,EVAL_GRP_DESCR,ESTN_UNIT_CN,TOTAL_AREA,N) |>
  summarise(estimate=sum(YSUM_HD*EXPNS),
            total_plots=unique(N),
            domain_plots=sum(NUMBER_PLOTS_IN_DOMAIN,na.rm=TRUE),
            non_zero_plots=sum(NON_ZERO_PLOTS,na.rm=TRUE),
            var_of_estimate = unique(TOTAL_AREA) * unique(TOTAL_AREA) / total_plots*
               ((sum(W_H * N_H *
                     (((YSUM_HD_SQR / N_H) -
                     ((YSUM_HD / N_H) *
                     ((YSUM_HD / N_H)))) /
                     (N_H - 1)))) +
               1 / total_plots *
               (sum((1 - W_H) * N_H *
                     (((YSUM_HD_SQR / N_H) -
                     ((YSUM_HD / N_H) *
                     (YSUM_HD / N_H))) /
                     (N_H - 1))))))
  
```

