---
title: "FIA-Post-Hurricane-Assessment"
author: "Ben Branoff & Andrés Baeza-Castro"
date: "2026-08-10"
output: github_document
always_allow_html: true
---

# FIA-Post-Hurricane-Assessment
Hurricane damage/loss estimates from collected plots after storms


Start by loading the plots from nims_srs_db.   Here, using the DBI library, which requires some initial setup for successful credentialed connections to the db.
Once setup, can connect to the "fiadb01p" data base and access any tables you are authorized to access. Then, we can use SQL queries passed as string 
arguments to retrieve the desired data. Here we are asking for the Florida (132024, 132025) and Georgia (122024, 122025) evaluations since hurricane Helene.
For now, we are only asking for the plots, their conditions, and their associated strata variables. Their trees and reference tables will be joined later.  
```{r load-plots,eval=FALSE}
library(DBI)
myconn <- dbConnect(odbc::odbc(), "fiadb01p", timeout = 10) # function change w/DBI
vars = paste0("select p.cn,p.plot_status_cd,p.invyr,p.eval_grp,p.statecd,p.countycd,p.measyear,p.measmon,p.measday,p.lat,p.lon,",
                    "c.condid,c.cond_status_cd,c.owncd,c.fortypcd,c.condprop_unadj,c.prop_basis,",
              "ps.evalid,ps.estn_unit_cn,ps.adj_factor_macr,ps.adj_factor_subp,ps.expns,",
              "pps.stratum_cn ")
tabs = "from fs_nims_fiadb_srs.plotsnap p,fs_nims_fiadb_srs.cond c,fs_nims_fiadb_srs.pop_stratum ps,fs_nims_fiadb_srs.pop_plot_stratum_assgn pps "
filts = paste0("where p.eval_grp in (132024,122024,132025,122025) 
and ps.evalid in(122501,122401,132501,132401)
and c.plt_cn = p.cn
and pps.stratum_cn = ps.cn
and pps.plt_cn = p.cn
and c.cond_status_cd = 1 and c.condprop_unadj IS NOT NULL")
#and to_date(p.measyear || '-' || LPAD(p.measmon, 2, '0') || '-' || LPAD(p.measday, 2, '0')
#               default null on conversion error,
#               'YYYY-MM-DD') >= date '2024-09-26'") 
plots = dbGetQuery(myconn,paste0(vars,tabs,filts))
plots
```
```{r load-plots-hidden,eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE}
library(Cyclones)
library(terra)
library(sf)
library(dplyr)
library(knitr)
library(kableExtra)
#write.csv(plots,"E:/OneDrive - USDA/Hurricanes/Followup Assessment/plotsHelene.csv")
plots <- read.csv("E:/OneDrive - USDA/Hurricanes/Followup Assessment/plotsHelene.csv")
plots_geo <- st_as_sf(plots,coords=c("LON","LAT"),crs=4326)
plots |> head() |>
  kable(format='html')|>
  kable_styling() |>
  scroll_box(width='100%')
```

The above is a small subset, but there are around 8000 plots that have been collected since the storm in Georgia and Florida.

Now, use the Cyclones package to recreate the winds for Helene and create wind intensity zones. For now, Cyclones is only available on
Github, so an extra package ("remotes") is required to install it. 

First, get the storm track information. This is the foundation for being able to retreive the winds (and rain and storm surge).

```{r load-winds,eval=FALSE}
##  install
install.packages('remotes')
remotes::install_github("BBranoff/Cyclones")
library(Cyclones)
library(terra)
library(sf)
library(dplyr)
##  we will use the rnaturalearth package to get state boundaries
usa <- rnaturalearth::ne_countries(country="united states of america",scale=10)
##  can also download the county boundaries
counties <- read_sf("/vsizip//vsicurl/https://naciscdn.org/naturalearth/10m/cultural/ne_10m_admin_2_counties.zip",
                        layer = "ne_10m_admin_2_counties")
##  also, convert the plots to a spatial object
plots_geo <- st_as_sf(plots,coords=c("LON","LAT"),crs=4326)
##  get the storm information
Helene <- get_storms(name="HELENE",season=2024)
Helene
```
```{r load-winds-hidden,eval=TRUE,echo=FALSE}
#saveRDS(list(storm=Helene,extent=Helene_extents,winds=wrap(Helene_winds)),"E:/OneDrive - USDA/Hurricanes/Followup Assessment/Helene.RDS")
HeleneStuff <- readRDS("E:/OneDrive - USDA/Hurricanes/Followup Assessment/Helene.RDS")
Helene <- HeleneStuff$storm
Helene_extents <- HeleneStuff$extent
Helene_winds <- lapply(HeleneStuff$winds,unwrap)
usa <- rnaturalearth::ne_countries(country="united states of america",scale=10)
counties <- read_sf("/vsizip//vsicurl/https://naciscdn.org/naturalearth/10m/cultural/ne_10m_admin_2_counties.zip",
                        layer = "ne_10m_admin_2_counties")
```

Now, calculate the wind field. Here we just use a theoretical (Boose) equation based on the maximum wind speed.
We calulate the wind onto a grid with a spatial resolution of 5km.

```{r winds,eval=FALSE}
###  first, interpolate the spatial wind information to every 30 mins
Helene_extents <- make_extents(Helene,t_res=30)
###  then interpolate spatially as a raster
Helene_winds <- get_wind(Helene_extents,method="Boose",agg=TRUE,s_res=5000)
###  classify the maximum wind speed into the Saffir-Simpson scale
Helene_Cat <- as.factor(classify(Helene_winds$boose$max_msw,
                       matrix(c(0,33,0,
                                33,43,1,
                                43,50,2,
                                50,60,3,
                                60,70,4,
                                70,Inf,5),ncol=3,byrow=TRUE),include.lowest=TRUE))
plot(Helene_Cat)
plot(st_geometry(st_transform(usa,crs(Helene_Cat))),add=TRUE)
```
```{r winds-hidden,eval=TRUE,echo=FALSE,fig.width=9,fig.height=5,fig.cap = "Fig 1. Hurricane Helene's wind field, categorized into intensity zones." }
Helene_Cat <- as.factor(classify(Helene_winds$boose$max_msw,
                       matrix(c(0,33,0,
                                33,43,1,
                                43,50,2,
                                50,60,3,
                                60,70,4,
                                70,Inf,5),ncol=3,byrow=TRUE),include.lowest=TRUE))
plot(Helene_Cat)
plot(st_geometry(st_transform(usa,crs(Helene_Cat))),add=TRUE)
```

With the hurricane intensity areas defined, we can assign them to each plot. We also need to define the area of each wind zone. 
These will be our artificial 'strata', and we will need to know the total area in order to calculate the expansion factors. 

```{r extract,eval=TRUE,fig.width=9,fig.height=5,fig.cap = "Fig 2. Hurricane Helene's wind zones with the plots overlayed. Plots are assigned their respective wind zones and their expansion factors come from the total area of each wind zone divided by the number of plots in each zone." }
###  first, mask out the areas outside of our interest
###  for now, this can just be the plots
###  but this probably needs to be 'forested' areas
Helene_Cat_plots <- trim(mask(Helene_Cat,st_transform(st_as_sf(st_concave_hull(st_union(plots_geo),ratio=0.02)),crs(Helene_Cat))))
##  now calculate area of each wind zone
##  important to do this while the raster is in projected, not geographic, coordinates
Areas <- table(values(Helene_Cat_plots ))
##  this is the number of pixels of each wind zone
##  multiply by the raster resolution to get the area, in square meters
Areas <- Areas*res(Helene_Cat_plots )[1]*res(Helene_Cat_plots )[2]
##  then convert to acres
Areas <- Areas/4046.86

###  project onto geographic coordinates
Helene_Cat_plots <- project(Helene_Cat_plots,"epsg:4326")
plot(Helene_Cat_plots)
###  overlay the plots
plot(st_geometry(st_as_sf(plots_geo,coords=c("LON","LAT"),crs=4326)),add=TRUE,pch=19,cex=0.5)
plot(st_geometry(usa),add=TRUE,border="darkgrey")
plot(st_geometry(counties),add=TRUE,border="darkgrey")
```

We now have the plots and the areas associated with each wind zone. From here, we can begin to assemble the
necessary tables and calculate the desired estimates. 

Beginning with assembling the necessary tables. Forested area is likely the most straight forward. We will get the 
forested acreage of each plot and, using a custom expansion factor, estimate the total forested acreage in each wind zone.
```{r evalidator_f,eval=TRUE}
fiadb_api_GET = function(url){

  # make request
  resp <- httr::GET(url=url)
  # parse response from JSON to R list
  respObj <- httr::content(resp, "parsed", encoding = "ISO-8859-1")

  # create empty output list
  outputList = list()

  # add estimates data frame to output list
  if (! is.null(respObj$estimates)) {
      outputList$estimates <- format_estimate(respObj$estimates)
  } else  {
      print("Problem with URL or API. No estimate returned.")
      return(list())
  }


  #Use lapply to break apart subtotal list, then call sapply to format with names
  if (! is.null(respObj$subtotals))
        outputList$subtotals <- sapply(lapply(respObj$subtotals, "["),
                                       format_estimate, simplify = F, USE.NAMES = T)

  # totals data frame
  if (! is.null(respObj$totals))
    outputList$totals <- format_estimate(respObj$totals)


  # add estimate metadata, doesn't need to be reformatted
  if (! is.null(respObj$metadata)) outputList$metadata <- respObj$metadata

  return(outputList)
}
format_estimate = function(respList) {

  return(as.data.frame(do.call(rbind, respList)))

}
```
```{r evalidate,eval=TRUE}
###  first, join the total 'strata' (wind zone) areas to the plot table
plots_geo$WindZone <- extract(Helene_Cat_plots,plots_geo)$max_msw
plots_geo<- plots_geo |>
  left_join(as.data.frame(Areas)|>rename(Acres=Freq),by=join_by(WindZone==Var1))|>
  group_by(WindZone)|>
  mutate(EXPNS_new = Acres/n())

##  first testing the concept on the full evaluation with the original expansion factors
ForestedAcres <- plots_geo |>
  mutate(ForestedAcres = if_else(PROP_BASIS=="MACR",
                                 CONDPROP_UNADJ*ADJ_FACTOR_MACR,CONDPROP_UNADJ*ADJ_FACTOR_SUBP))|>
  filter(EVALID %in% c(122401,132401),EVAL_GRP %in% c(122024,132024))|>
  group_by(EVAL_GRP)|>
  summarise(ForestedAcres=sum(ForestedAcres*EXPNS))|>
  st_drop_geometry()
###  our custom estimate
ForestedAcres
###  EVALIDATOR for Florida
fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=122024&outputFormat=NJSON")$totals$ESTIMATE
###  EVALIDATOR for Georgia
fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=132024&outputFormat=NJSON")$totals$ESTIMATE

```

From the above, the DIY ad-hoc estimate matches the total forested acres for these two evaluations from EVALIDATOR. To apply it to our 
custom stata (wind zones), we only need to change the EXPNS from the full evaluation tables to our custom EXPNS_new that we calculated
based on the calculated area of the strata and the number of plots in each strata. 

We also add in the common request that this be done by county and by forest type, although we limit the forest type to only the most common
requested types. 
```{r forestedarea,eval=TRUE}
##  first testing the concept on the full evaluation with the original expansion factors
HeleneAcres <- plots_geo |>st_drop_geometry()|>
  mutate(ForestedAcres = if_else(PROP_BASIS=="MACR",
                                 CONDPROP_UNADJ*ADJ_FACTOR_MACR,CONDPROP_UNADJ*ADJ_FACTOR_SUBP))|>
  #filter(EVALID %in% c(122401,132401),EVAL_GRP %in% c(122024,132024),
  #       (FORTYPCD>=140&FORTYPCD<=150)|(FORTYPCD>=160&FORTYPCD<=170)|(FORTYPCD>=400&FORTYPCD<=500)|(FORTYPCD>=970&FORTYPCD<=980))|>  #limit to common forest types
    mutate(FORTYP = if_else(FORTYPCD>=140&FORTYPCD<=150,"longleaf/slash pine",
                          if_else(FORTYPCD>=160&FORTYPCD<=170,"loblolly/shortleaf pine",
                                  if_else(FORTYPCD>=400&FORTYPCD<=500,"oak/pine",
                                          if_else(FORTYPCD>=970&FORTYPCD<=980,"woodland hardwoods","all others")))))|>
  left_join(counties|>filter(REGION_COD %in% c(12,13))|>mutate(COUNTYCD=as.numeric(substr(CODE_LOCAL,3,6)),
                                                               REGION_COD=as.numeric(REGION_COD))|>
              select(NAME,REGION_COD,COUNTYCD)|>rename(County=NAME),by=join_by(STATECD==REGION_COD,COUNTYCD))
  
HeleneAcres <- rbind(HeleneAcres,
                     HeleneAcres|>mutate(FORTYP="All",County="All"),
                     HeleneAcres|>mutate(FORTYP="All"),
                     HeleneAcres|>mutate(County="All"))|>
  group_by(STATECD,EVAL_GRP,FORTYP,County,WindZone)|>
  summarise(nplots=n(),
            ForestedAcres=sum(ForestedAcres*EXPNS_new))
  
HeleneAcres|>filter(EVAL_GRP==122024,FORTYP=="All",County=="All")|>pull(ForestedAcres)|>sum(na.rm=TRUE)
fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=122024&outputFormat=NJSON")$totals$ESTIMATE
```

The new total acreage does not match the total acreage from EVALIDATOR. There are 55 plots that are not in any wind zone, which is part of the reason but
the total is about half the EVALIDATOR total, which suggests there is something else missing. 