---
title: "FIA-Post-Hurricane-Assessment"
author: "Ben Branoff & Andrés Baeza-Castro"
date: "2026-08-10"
output: github_document
always_allow_html: true
---

# FIA-Post-Hurricane-Assessment
Hurricane damage/loss estimates from collected plots after storms


Start by loading the plots from nims_srs_db.   Here, using the DBI library, which requires some initial setup for successful credentialed connections to the db.
Once setup, can connect to the "fiadb01p" data base and access any tables you are authorized to access. Then, we can use SQL queries passed as string 
arguments to retrieve the desired data. Here we are asking for the Florida (132024, 132025) and Georgia (122024, 122025) evaluations since hurricane Helene.
For now, we are only asking for the plots, their conditions, and their associated strata variables. Their trees and reference tables will be joined later.  
```{r load-plots,eval=FALSE}
library(DBI)
myconn <- dbConnect(odbc::odbc(), "fiadb01p", timeout = 10) # function change w/DBI
vars = paste0("select p.cn,p.prev_plt_cn,p.plot_status_cd,p.invyr,p.eval_grp,p.statecd,p.countycd,p.measyear,p.measmon,p.measday,p.lat,p.lon,",
                    "c.condid,c.cond_status_cd,c.owncd,c.fortypcd,c.condprop_unadj,c.prop_basis,",
              "ps.evalid,ps.estn_unit_cn,ps.adj_factor_macr,ps.adj_factor_subp,ps.expns,",
              "pps.stratum_cn ")
tabs = "from fs_nims_fiadb_srs.plotsnap p,fs_nims_fiadb_srs.cond c,fs_nims_fiadb_srs.pop_stratum ps,fs_nims_fiadb_srs.pop_plot_stratum_assgn pps "
filts = paste0("where p.eval_grp in (132024,122024,132025,122025) 
and ps.evalid in(122501,122401,132501,132401)
and c.plt_cn = p.cn
and pps.stratum_cn = ps.cn
and pps.plt_cn = p.cn
and c.cond_status_cd = 1 and c.condprop_unadj IS NOT NULL")
#and to_date(p.measyear || '-' || LPAD(p.measmon, 2, '0') || '-' || LPAD(p.measday, 2, '0')
#               default null on conversion error,
#               'YYYY-MM-DD') >= date '2024-09-26'") 
plots = dbGetQuery(myconn,paste0(vars,tabs,filts))
plots
```
```{r load-plots-hidden,eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE}
library(Cyclones)
library(terra)
library(sf)
library(dplyr)
library(knitr)
library(kableExtra)
#write.csv(plots,"E:/OneDrive - USDA/Hurricanes/Followup Assessment/plotsHelene.csv")
plots <- read.csv("E:/OneDrive - USDA/Hurricanes/Followup Assessment/plotsHelene.csv")
plots_geo <- st_as_sf(plots,coords=c("LON","LAT"),crs=4326)
plots |> head() |>
  kable(format='html')|>
  kable_styling() |>
  scroll_box(width='100%')
```

The above is a small subset, but there are around 8000 plots that have been collected since the storm in Georgia and Florida.

Now, use the Cyclones package to recreate the winds for Helene and create wind intensity zones. For now, Cyclones is only available on
Github, so an extra package ("remotes") is required to install it. For this workflow, the hurricane wind information is provided as part of the download and there is no need to recalculate, but the process is demonstrated.
```{r download-winds,eval=TRUE}
HeleneStuff <- readRDS("E:/OneDrive - USDA/Hurricanes/Followup Assessment/FIA-Post-Hurricane-Assessment/data/Helene.RDS")
Helene <- HeleneStuff$storm
Helene_extents <- HeleneStuff$extent
Helene_winds <- lapply(HeleneStuff$winds,unwrap)
usa <- rnaturalearth::ne_countries(country="united states of america",scale=10)
counties <- read_sf("/vsizip//vsicurl/https://naciscdn.org/naturalearth/10m/cultural/ne_10m_admin_2_counties.zip",
                        layer = "ne_10m_admin_2_counties")
```


First, get the storm track information. This is the foundation for being able to retrieve the winds (and rain and storm surge).
```{r load-winds,eval=FALSE}
##  install
install.packages('remotes')
remotes::install_github("BBranoff/Cyclones")
library(Cyclones)
library(terra)
library(sf)
library(dplyr)
##  we will use the rnaturalearth package to get state boundaries
usa <- rnaturalearth::ne_countries(country="united states of america",scale=10)
##  can also download the county boundaries
counties <- read_sf("/vsizip//vsicurl/https://naciscdn.org/naturalearth/10m/cultural/ne_10m_admin_2_counties.zip",
                        layer = "ne_10m_admin_2_counties")
##  also, convert the plots to a spatial object
plots_geo <- st_as_sf(plots,coords=c("LON","LAT"),crs=4326)
##  get the storm information
Helene <- get_storms(name="HELENE",season=2024)
Helene
```


Now, calculate the wind field. Here we just use a theoretical (Boose) equation based on the maximum wind speed.
We calculate the wind onto a grid with a spatial resolution of 5km.

```{r winds,eval=FALSE}
###  first, interpolate the spatial wind information to every 30 mins
Helene_extents <- make_extents(Helene,t_res=30)
###  then interpolate spatially as a raster
Helene_winds <- get_wind(Helene_extents,method="Boose",agg=TRUE,s_res=5000)
###  classify the maximum wind speed into the Saffir-Simpson scale
Helene_Cat <- as.factor(classify(Helene_winds$boose$max_msw,
                       matrix(c(0,33,0,
                                33,43,1,
                                43,50,2,
                                50,60,3,
                                60,70,4,
                                70,Inf,5),ncol=3,byrow=TRUE),include.lowest=TRUE))
plot(Helene_Cat)
plot(st_geometry(st_transform(usa,crs(Helene_Cat))),add=TRUE)
```
```{r winds-hidden,eval=TRUE,echo=FALSE,fig.width=9,fig.height=5,fig.cap = "Fig 1. Hurricane Helene's wind field, categorized into intensity zones." }
Helene_Cat <- as.factor(classify(Helene_winds$boose$max_msw,
                       matrix(c(0,33,0,
                                33,43,1,
                                43,50,2,
                                50,60,3,
                                60,70,4,
                                70,Inf,5),ncol=3,byrow=TRUE),include.lowest=TRUE))
plot(Helene_Cat)
plot(st_geometry(st_transform(usa,crs(Helene_Cat))),add=TRUE)
```

With the hurricane intensity areas defined, we can assign them to each plot. We also need to define the area of each wind zone. 
These will be our artificial 'strata', and we will need to know the total area in order to calculate the expansion factors. 

```{r extract,eval=TRUE,fig.width=9,fig.height=9,fig.cap = "Fig 2. Hurricane Helene's wind zones with the plots overlayed. Plots are assigned their respective wind zones and their expansion factors come from the total area of each wind zone divided by the number of plots in each zone." }
###  first, mask out the areas outside of our interest
###  for now, this can just be the plots
###  but this probably needs to be 'forested' areas
Helene_Cat_plots <- trim(mask(Helene_Cat,st_transform(st_as_sf(st_concave_hull(st_union(plots_geo),ratio=0.001)),crs(Helene_Cat))))
##  now calculate area of each wind zone
##  important to do this while the raster is in projected, not geographic, coordinates
##  also important to do by state, as this is how the evaluation is done
Areas12 <- table(values(mask(Helene_Cat_plots,st_transform(counties|>filter(REGION_COD==12),crs(Helene_Cat_plots)))))
Areas13 <- table(values(mask(Helene_Cat_plots,st_transform(counties|>filter(REGION_COD==13),crs(Helene_Cat_plots)))))
Areas <- rbind(data.frame(Areas12)|>mutate(STATECD=12),
               data.frame(Areas13)|>mutate(STATECD=13))
##  this is the number of pixels of each wind zone
##  multiply by the raster resolution to get the area, in square meters
Areas$Freq <- Areas$Freq*res(Helene_Cat_plots )[1]*res(Helene_Cat_plots )[2]
##  then convert to acres
Areas$Acres <- Areas$Freq/4046.86

###  project onto geographic coordinates
Helene_Cat_plots <- project(Helene_Cat_plots,"epsg:4326")
plot(Helene_Cat_plots)
###  overlay the plots
plot(st_geometry(st_as_sf(plots_geo,coords=c("LON","LAT"),crs=4326)),add=TRUE,pch=19,cex=0.01)
plot(st_geometry(usa),add=TRUE,border="darkgrey")
plot(st_geometry(counties),add=TRUE,border="darkgrey")
```

We now have the plots and the areas associated with each wind zone. From here, we can begin to assemble the
necessary tables and calculate the desired estimates. 

Beginning with assembling the necessary tables. Forested area is likely the most straight forward. We will get the 
forested acreage of each plot and, using a custom expansion factor, estimate the total forested acreage in each wind zone.
```{r evalidator_f,eval=TRUE,echo=FALSE}
fiadb_api_GET = function(url){

  # make request
  resp <- httr::GET(url=url)
  # parse response from JSON to R list
  respObj <- httr::content(resp, "parsed", encoding = "ISO-8859-1")

  # create empty output list
  outputList = list()

  # add estimates data frame to output list
  if (! is.null(respObj$estimates)) {
      outputList$estimates <- format_estimate(respObj$estimates)
  } else  {
      print("Problem with URL or API. No estimate returned.")
      return(list())
  }


  #Use lapply to break apart subtotal list, then call sapply to format with names
  if (! is.null(respObj$subtotals))
        outputList$subtotals <- sapply(lapply(respObj$subtotals, "["),
                                       format_estimate, simplify = F, USE.NAMES = T)

  # totals data frame
  if (! is.null(respObj$totals))
    outputList$totals <- format_estimate(respObj$totals)


  # add estimate metadata, doesn't need to be reformatted
  if (! is.null(respObj$metadata)) outputList$metadata <- respObj$metadata

  return(outputList)
}
format_estimate = function(respList) {

  return(as.data.frame(do.call(rbind, respList)))

}
```
```{r evalidate,eval=TRUE}
###  first, join the total 'strata' (wind zone) areas to the plot table
plots_geo$WindZone <- extract(Helene_Cat_plots,plots_geo)$max_msw
plots_geo<- plots_geo |>
  left_join(Areas,by=join_by(WindZone==Var1,STATECD))|>
  group_by(WindZone)|>
  mutate(EXPNS_new = Acres/length(unique(CN)))

##  first testing the concept on the full evaluation with the original expansion factors
ForestedAcres <- plots_geo |>
  mutate(ForestedAcres = if_else(PROP_BASIS=="MACR",
                                 CONDPROP_UNADJ*ADJ_FACTOR_MACR,CONDPROP_UNADJ*ADJ_FACTOR_SUBP))|>
  filter(EVALID %in% c(122401,132401),EVAL_GRP %in% c(122024,132024))|>
  group_by(EVAL_GRP)|>
  summarise(ForestedAcres=sum(ForestedAcres*EXPNS))|>
  st_drop_geometry()
###  our custom estimate versus EVALIDATOR
ForestedAcres |>
  mutate(EVALIDATORacres = as.numeric(c(fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=122024&outputFormat=NJSON")$totals$ESTIMATE,
                             fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=132024&outputFormat=NJSON")$totals$ESTIMATE)))

```

From the above, the DIY ad-hoc estimate matches the total forested acres for these two evaluations from EVALIDATOR. To apply it to our 
custom stata (wind zones), we only need to change the EXPNS from the full evaluation tables to our custom EXPNS_new that we calculated
based on the calculated area of the strata and the number of plots in each strata. 

We also add in the common request that this be done by county and by forest type, although we limit the forest type to only the most common
requested types. 
```{r forestedarea,eval=TRUE,message=FALSE}
##  first testing the concept on the full evaluation with the original expansion factors
HeleneAcres <- plots_geo |>st_drop_geometry()|>
  mutate(ForestedAcres = if_else(PROP_BASIS=="MACR",
                                 CONDPROP_UNADJ*ADJ_FACTOR_MACR,CONDPROP_UNADJ*ADJ_FACTOR_SUBP))|>
  #filter(EVALID %in% c(122401,132401),EVAL_GRP %in% c(122024,132024),
  #       (FORTYPCD>=140&FORTYPCD<=150)|(FORTYPCD>=160&FORTYPCD<=170)|(FORTYPCD>=400&FORTYPCD<=500)|(FORTYPCD>=970&FORTYPCD<=980))|>  #limit to common forest types
    mutate(FORTYP = if_else(FORTYPCD>=140&FORTYPCD<=150,"longleaf/slash pine",
                          if_else(FORTYPCD>=160&FORTYPCD<=170,"loblolly/shortleaf pine",
                                  if_else(FORTYPCD>=400&FORTYPCD<=500,"oak/pine",
                                          if_else(FORTYPCD>=970&FORTYPCD<=980,"woodland hardwoods","all others")))))|>
  left_join(counties|>filter(REGION_COD %in% c(12,13))|>mutate(COUNTYCD=as.numeric(substr(CODE_LOCAL,3,6)),
                                                               REGION_COD=as.numeric(REGION_COD))|>
              select(NAME,REGION_COD,COUNTYCD)|>rename(County=NAME),by=join_by(STATECD==REGION_COD,COUNTYCD))
  
HeleneAcres <- rbind(HeleneAcres,
                     HeleneAcres|>mutate(FORTYP="All",County="All"),
                     HeleneAcres|>mutate(FORTYP="All"),
                     HeleneAcres|>mutate(County="All"))|>
  group_by(STATECD,EVAL_GRP,FORTYP,County,WindZone)|>
  summarise(nplots=length(unique(CN)),
            ForestedAcres=sum(ForestedAcres*EXPNS_new))
  
HeleneAcres|>
  filter(EVAL_GRP%in% c(122024,132024),FORTYP=="All",County=="All")|>
  group_by(EVAL_GRP) |>
  summarise(ForestedAcres=sum(ForestedAcres,na.rm=TRUE))|>
  mutate(EVALIDATORSacres = as.numeric(c(fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=122024&outputFormat=NJSON")$totals$ESTIMATE,
                                                                                                                                                                                        fiadb_api_GET(url="https://apps.fs.usda.gov/fiadb-api/fullreport?snum=02&wc=132024&outputFormat=NJSON")$totals$ESTIMATE)))
```

These estimates are close to the totals from EVALIDATOR but a little off. This is probably because we are using the total area of the strata to calculate the expansion factor instead of using only the forested area inside the strata (?).


To do tree level estimates (volume, biomass, etc.) we need the tree tables as well as the GRM tables to back-calculate mortality volume or biomass for when the tree died. 

first the GRM, as this is probably the most complex.
```{r load-plots,eval=FALSE}
library(DBI)
myconn <- dbConnect(odbc::odbc(), "fiadb01p", timeout = 10) # function change w/DBI

sql <- "select SUM(GRM.TPAMORT_UNADJ) as estimatedvalue from (select * from fs_nims_fiadb_srs.pop_stratum t where evalid = 132503) POP_STRATUM join (select stratum_cn, count(cn) as p2pointcnt_sy from fs_nims_fiadb_srs.POP_PLOT_STRATUM_ASSGN t where evalid = 132503 and invyr = 2025 group by stratum_cn) PPSA on cn = PPSA.stratum_cn join (select * from fs_nims_fiadb_srs.POP_ESTN_UNIT t where evalid = 132503) PEU on PEU.CN = POP_STRATUM.ESTN_UNIT_CN join fs_nims_fiadb_srs.POP_PLOT_STRATUM_ASSGN POP_PLOT_STRATUM_ASSGN ON (POP_PLOT_STRATUM_ASSGN.STRATUM_CN = POP_STRATUM.CN) JOIN fs_nims_fiadb_srs.plotsnap PLOT ON (POP_PLOT_STRATUM_ASSGN.PLT_CN = PLOT.CN) JOIN fs_nims_fiadb_srs.COND COND ON (COND.PLT_CN = PLOT.CN) JOIN (SELECT P.PREV_PLT_CN, T.* FROM fs_nims_fiadb_srs.plotsnap P JOIN fs_nims_fiadb_srs.TREE_VW T ON (P.CN = T.PLT_CN)) TREE ON ((TREE.CONDID = COND.CONDID) AND (TREE.PLT_CN = COND.PLT_CN)) LEFT OUTER JOIN fs_nims_fiadb_srs.plotsnap PPLOT ON (PLOT.PREV_PLT_CN = PPLOT.CN) LEFT OUTER JOIN fs_nims_fiadb_srs.COND PCOND ON ((TREE.PREVCOND = PCOND.CONDID) AND (TREE.PREV_PLT_CN = PCOND.PLT_CN)) LEFT OUTER JOIN fs_nims_fiadb_srs.TREE_VW PTREE ON (TREE.PREV_TRE_CN = PTREE.CN) LEFT OUTER JOIN fs_nims_fiadb_srs.TREE_GRM_BEGIN TRE_BEGIN ON (TREE.CN = TRE_BEGIN.TRE_CN) LEFT OUTER JOIN fs_nims_fiadb_srs.TREE_GRM_MIDPT TRE_MIDPT ON (TREE.CN = TRE_MIDPT.TRE_CN) LEFT OUTER JOIN (SELECT TRE_CN,DIA_BEGIN,DIA_MIDPT,DIA_END,MICR_COMPONENT_AL_TIMBER AS COMPONENT,MICR_SUBPTYP_GRM_AL_TIMBER AS SUBPTYP_GRM,MICR_TPAMORT_UNADJ_AL_TIMBER AS TPAMORT_UNADJ FROM fs_nims_fiadb_srs.TREE_GRM_COMPONENT) GRM ON (TREE.CN = GRM.TRE_CN) JOIN fs_nims_fiadb_srs.REF_SPECIES ON (TREE.SPCD = REF_SPECIES.SPCD) WHERE REF_SPECIES.WOODLAND = 'N' AND POP_PLOT_STRATUM_ASSGN.Invyr = 2025"

sql <- "select sum(POP_STRATUM.EXPNS) as estimatedvalue from (select * from fs_nims_fiadb_srs.pop_stratum t where evalid = 132503) POP_STRATUM"
test <- dbGetQuery(myconn,sql)


JOIN FS_nims_srs.REF_SPECIES
ON (TREE.SPCD = REF_SPECIES.SPCD)
WHERE REF_SPECIES.WOODLAND = 'N'
AND POP_PLOT_STRATUM_ASSGN.Invyr = 2025
AND 1 = 1
"





trees <- dbGetQuery(myconn,"select t.*,p.* from fs_nims_fiadb_srs.tree_vw t,fs_nims_fiadb_srs.plotsnap p where p.cn=t.plt_cn and p.eval_grp in (122024,132024,122025,132025)")
#write.csv(trees,"E:/OneDrive - USDA/Hurricanes/Followup Assessment/plotsHelene.csv")
trees <- read.csv("E:/OneDrive - USDA/Hurricanes/Followup Assessment/plotsHelene.csv")
trees <- dbGetQuery(myconn,"select * from fs_nims_fiadb_srs.pop_stratum ps where evalid in (132503,122503)") |>
  left_join(dbGetQuery(myconn,"select * from fs_nims_fiadb_srs.pop_plot_stratum_assgn where evalid in (122503,132503) and invyr=2025")|>
              group_by(STRATUM_CN)|>
              mutate(p2pointcnt_sy=length(unique(CN))),by=join_by(CN==STRATUM_CN)) |>
  left_join(dbGetQuery(myconn,"select * from fs_nims_fiadb_srs.pop_estn_unit where evalid in (122503,132503)"),by=join_by(ESTN_UNIT_CN==CN))|>
  left_join(plots|>mutate(CN=as.character(CN)),by=join_by(PLT_CN==CN))|>
  left_join(dbGetQuery(myconn,"select * from fs_nims_fiadb_srs.tree_vw where statecd in (12,13) and invyr=2025"),by=join_by(PLT_CN,CONDID))
trees <- trees |>
  left_join(plot,by=join_by(PREV_PLOT_CN==PLT_CN))


JOIN FS_nims_srs.NIMS_COND_VW COND
ON (COND.PLT_CN = PLOT.CN)
JOIN (SELECT P.PREV_PLT_CN, T.*
        FROM FS_nims_srs.NIMS_PLOT_VW P
      JOIN FS_nims_srs.NIMS_TREE_VW T
      ON (P.CN = T.PLT_CN)) TREE
ON ((TREE.CONDID = COND.CONDID) AND (TREE.PLT_CN = COND.PLT_CN))
###  same as
#pop_plot_stratum_assgn2 = dbGetQuery(myconn,"select * from fs_nims_fiadb_srs.pop_plot_stratum_assgn where evalid=132503 and invyr=2025")|>
# group_by(STRATUM_CN)|>
#  summarise(p2pointcnt_sy=length(unique(CN)))
ppsa <- pop_stratum |>
  left_join(pop_plot_stratum_assgn,by=join_by(CN==STRATUM_CN))
peu <- dbGetQuery(myconn,"select * from fs_nims_fiadb_srs.pop_estn_unit where evalid = 132503")


vars = paste0("select p.cn,p.plot_status_cd,p.invyr,p.eval_grp,p.statecd,p.countycd,p.measyear,p.measmon,p.measday,p.lat,p.lon,",
                    "c.condid,c.cond_status_cd,c.owncd,c.fortypcd,c.condprop_unadj,c.prop_basis,",
              "ps.evalid,ps.estn_unit_cn,ps.adj_factor_macr,ps.adj_factor_subp,ps.expns,",
              "pps.stratum_cn ")
tabs = "from fs_nims_srs.nims_pop_stratum ps, fs_nims_srs.pop_plot_stratum_assgn pps  fs_nims_srs.nims_plot_vw pplot,,fs_nims_srs.nims_tree_vw ptree,fs_nims_srs.nims_tree_grm_begin tre_begin,fs_nims_srs.nims_tree_grm_midpt tre_midpt,  "
filts = paste0("where p.eval_grp in (132024,122024,132025,122025) 
and ps.evalid in(122501,122401,132501,132401)
and c.plt_cn = p.cn
and pps.stratum_cn = ps.cn
and pps.plt_cn = p.cn
and c.cond_status_cd = 1 and c.condprop_unadj IS NOT NULL")

plots = dbGetQuery(myconn,paste0(vars,tabs,filts))
plots
```
